---
title: "readme"
author: "Lissa Breugelmans"
date: "2024-10-15"
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# pdindicatoR

Phylogenetic diversity (PD) is a measure of biodiversity which takes evolution into account. It is calculated as the sum of the lengths of the phylogenetic tree branches representing the minimum tree-spanning path among a group of species.
Phylogenetic diversity can be used in conservation planning to maximise a variety of features, meaning we do not aim to conserve specific features, but rather want to boost a diverse range of features. Conserving a variety of features could be particularly useful in light of the changing environmental conditions, as we can only guess which features will be important in the future.

In this package, we provide a workflow to calculate a metric that gives information about how well PD of a certain higher taxonomic group is currently safeguarded by protected areas and a spatial visualisation which can be used to identify potential directions for future expansion of protected areas.


## Installation

pdindicatoR package can be loaded using devtools() from the Github repository.

## Example

This example shows a basic workflow for using the functions in the pdindicatoR package to calculate PD from a phylogenetic tree and an occurrence cube with occurrences for a certain higher taxon, produce a gridded map of PD scores with colour gradient scale, and show the overlap with protected areas. 

```{r loadlib, echo=T, results='hide', message=F, warning=F}
# Load packages
library(tidyverse)
library(rotl)
library(sf)
library(gdalUtilities)
library(ape)
library(rnaturalearth)
library(purrr)
```

### Editing the config file

In order to start the workflow, the user should edit config.R in order to specify the location of the phylogenetic tree (Newick format) and data cube (.csv format) files.

```{r}
tree_path <- "./data/Musteloidea_tree.tre"
cube_path <- "./data/cube_musteloidea_BE_databricks.csv"
```
### Loading tree and datacube

A phylogenetic tree and occurrence datacube for selected taxa should be provided by the user. 

The phylogenetic tree should be specified in Newick format and contain branch lengths. Published phylogenetic trees can be downloaded from <a href="https://tree.opentreeoflife.org/"> Open Tree of Life project</a>. The rotl package also can be used to query the OTL database and retrieve published trees. The following example code can be adapted to find published trees for a specific higher taxon.

```{r, echo=FALSE, results='hide'}
mono_id <- tnrs_match_names("Musteloideae") # Returns matching name and ott_id of taxon
studies_properties() # Returns list of properties that can be used in the search
furry_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName", value = "Mustelidae") # Search for unique name
furry_studies <- studies_find_studies(property = "ot:focalClade", value = 348043) # Search for ott_id
get_study_tree("pg_1939","tree3941") # Returns published tree
```

Occurrence datacubes can be generated by querying the GBIF SQL Occurrence Download API. <a href="https://techdocs.gbif.org/en/data-use/api-sql-downloads">This webpage</a> details how to query the API while <a href="https://techdocs.gbif.org/en/data-use/data-cubes">This webpage</a> contains instructions on how to write a SQL query that returns an occurrence cube. The API can also be queried using the <a href="https://docs.ropensci.org/rgbif/reference/occ_download_sql.html">occ_download_sql()</a> function of <a href="https://docs.ropensci.org/rgbif/index.html">rgbif</a>.

The phylogenetic tree and datacube are loaded into R and printed to confirm they are processed correctly.

```{r}
tree <- ape::read.tree(tree_path)
```
```{r}
plot(tree, cex=0.35,y.lim=75)
nodelabels(cex=0.35)
edgelabels(cex=0.35)
tiplabels(cex=0.35)
```
```{r}
cube <- read.csv(cube_path, stringsAsFactors = FALSE, sep=",")
# specify the correct seperator depending on the format of the datacube - occurrence datacubes downloaded through the GBIF SQL API  are currently tab seperated (sep="\t")
head(cube)
```
Please note that when using datacubes from another source, the column names are renamed to match the GBIF terms (see example datacube above).

### Load functions -- Can be removed once package is installable -- 

The pdindicatoR functions can be loaded using *source()*.
```{r}
# Load functions
source("./R/taxonmatch.R")
source("./R/append_ott_id.R")
source("./R/map_pd.R")
source("./R/pdindicator.R")
source("./R/convert_multipolygons.R")
source("./R/get_pd.R")
source("./R/calculate_faithpd.R")
source("./R/check_completeness.R")
source("./R/aggregate_cube.R")
```

### Matching species in phylogenetic tree and datacube

The leaf labels of a phylogenetic tree downloaded from the OTL database are specified as either species names or OTL id's (ott_id). We can use the function taxonmatch() to retrieve the corresponding GBIF id's.

```{r}
matched <- taxonmatch(tree)
head(matched)
```

Carefully evaluate the table with matches to ensure that matching scores are acceptable and that most species have a corresponding gbif_id. Species that cannot be reliable matched or that don't have an associated gbif_id, can not contribute to the PD calculation and should be removed.

```{r}
matched_nona <- matched %>% dplyr::filter(!is.na(gbif_id))
```

Then, we can use the function append_ott_id() to append the ott_id's as a new variable to the provided datacube, by joining on gbif_id.

```{r}
mcube <- append_ott_id(tree, cube, matched_nona)
head(mcube)
```

When species in the datacube are not included in the provided phylogenetic tree, the ott_id variable will be *NA*. We can use the function check_completeness() to see how complete the provided phylogenetic tree is.
```{r}
check_completeness(mcube)
```

Please note that occurrence records for species that are not part of the provided phylogenetic tree will need to be removed. In case this number is large, please consider searching for a more complete phylogenetic tree that covers all your species!

```{r}
mcube <- mcube %>% dplyr::filter(!is.na(ott_id))
```


### Calculate Phylogenetic Diversity for each grid cell

We first have to aggregate the occurrence cube in order to get a list of observed species for each grid cell. 

```{r}
aggr_cube <- aggregate_cube(mcube)
head(aggr_cube)
#print(aggr_cube[[294,4]])
```

We can then calculate the PD value for all grid cells by using the purrr:map function to apply the function get_PD() for each grid cell. The PD values will be appended to the datacube as a new column 'PD'.


```{r}
PD_cube <- aggr_cube %>% mutate(PD = unlist(purrr::map(orig_tiplabels, ~ get_pd(tree, unlist(.x)))))
head(PD_cube)
```


### Visualize PD on a map & calculate indicator

Finally we read in the EEA Grid shapefiles and merge them to the occurrence cube by joining on the eeaCellCode field. The PD cube can then be plotted and overlayn with a polygon layer depicting the boundaries of WDPA protected areas.

#### Reading in the grid shapefile

```{r}
grid_filepath <- "./shpfiles/EEA-reference-GRID-2013.gpkg"
# Use sf::st_layers(grid_filepath) to get an overview of the layers in the gpkg
grid <- sf::st_read(grid_filepath, layer = "be_1km_polygon")
# Use sf::st_crs(grid) to view the grid's CRS
```

#### Reading in the protected areas shapefile 

```{r}
pa_filepath <- "./shpfiles/20240528_protected_areas_BE.gpkg"
sf::st_layers(pa_filepath) # get an overview of the layers in the gpkg
pa <- sf::st_read(pa_filepath, layer = "NaturaSite_polygon")
# Make sure the CRS of this layer corresponds to the CRS of the used grid
# by using sf::st_crs(pa)
```

#### Plot PD map

The function map_pd() can be used to generate a map visualizing phylogenetic 
diversity for the geographic area that is used to generate the occurrence cube.
If more detailed maps are desired, the optional argument *bbox_custom* can be
used to delineate the bounding box. Coordinates for the desired geographic area 
can be determined using https://epsg.io/ and selecting the CRS of the used grid.

```{r}
PDmap <- map_pd(PD_cube, grid, "Musteloidea")
# Optionally specify a custom bounding box: bbox_custom <- c(xmin,xmax,ymin,ymax)
# PDmap <- map_pd(PD_cube, grid, "Musteloidea", bbox_custom)
```

