---
title: "readme"
author: "Lissa Breugelmans"
date: "2024-10-15"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# pdindicatoR


Phylogenetic diversity (PD) is a measure of biodiversity which takes evolution into account. It is calculated as the sum of the lengths of the phylogenetic tree branches representing the minimum tree-spanning path among a group of species.
Phylogenetic diversity can be used in conservation planning to maximise a variety of features, meaning we do not aim to conserve specific features, but rather want to boost a diverse range of features. Conserving a variety of features could be particularly useful in light of the changing environmental conditions, as we can only guess which features will be important in the future.

In this package, we provide a workflow to calculate a metric that gives information about how well PD of a certain higher taxonomic group is currently safeguarded by protected areas and a spatial visualisation which can be used to identify potential directions for future expansion of protected areas.

## Installation

pdindicatoR package can be loaded using devtools() from the Github repository.

## Example

This example shows a basic workflow for using the functions in the pdindicatoR package to calculate PD from a phylogenetic tree and an occurence cube with occurences for a certain higher taxon, produce a gridded map of PD scores with colour gradient scale, and show the overlap with protected areas. 

```{r loadlib, echo=T, results='hide', message=F, warning=F}
# Load packages
library(tidyverse)
library(rotl)
library(sf)
library(gdalUtilities)
library(ape)
library(rnaturalearth)
library(purrr)
```

### Editing the config file

In order to start the workflow, the user should edit config.R in order to specify the location of the phylogenetic tree (Newick format) and data cube (.csv format) files. In addition, a cutoff-value for 'high phylogenetic diversity' should be specified and the coordinates of the bounding box for the maps.

```{r}
tree_path <- "./data/Musteloidea_tree.tre"
cube_path <- "./data/cube_musteloidea_BE_databricks.csv"
```
### Loading tree and datacube

A phylogenetic tree and occurrence datacube for selected taxa should be provided by the user. The phylogenetic tree should be specified in Newick format and contain branch lengths. Published phylogenetic trees can be downloaded through the <a href="https://tree.opentreeoflife.org/"> Open Tree of Life project</a>. Occurrence datacubes can be generated by querying the GBIF SQL Occurrence Download API. <a href="https://techdocs.gbif.org/en/data-use/api-sql-downloads">This webpage</a> details how to query the API while <a href="https://techdocs.gbif.org/en/data-use/data-cubes">This webpage</a> contains instructions on how to write a SQL query that returns an occurrence cube. The API can also be queried using the <a href="https://docs.ropensci.org/rgbif/reference/occ_download_sql.html">occ_download_sql()</a> function of <a href="https://docs.ropensci.org/rgbif/index.html">rgbif</a>.

The phylogenetic tree and datacube are loaded into R and printed to confirm they are processed correctly.

```{r}
tree <- ape::read.tree(tree_path)
plot(tree, cex=0.45)
```
```{r}
cube <- read.csv(cube_path, stringsAsFactors = FALSE, sep=",") # specify the correct seperator depending on the format of the datacube - occurence datacubes downloaded through the GBIF SQL API  are currently tab seperated (sep="\t")
head(cube)
```
Please note that when using datacubes from another source, the column names are renamed to match the GBIF terms (see example datacube above).

### Load functions -- Can be removed once package is installable -- 

The pdindicatoR functions can be loaded using *source()*.
```{r}
# Load functions
source("./R/taxonmatch.R")
source("./R/append_ott_id.R")
source("./R/pdmap.R")
source("./R/pdindicator.R")
source("./R/convert_multipolygons.R")
source("./R/calculate_PD.R")
source("./R/check_completeness.R")
source("./R/aggregate_cube.R")
```

### Matching species in phylogenetic tree and datacube

The leaf labels of a phylogenetic tree downloaded from the OTL database are specified as either species names or OTL id's (ott_id). We can use the function taxonmatch() to retrieve the corresponding GBIF id's.

```{r}
matched <- taxonmatch(tree)
head(matched)
```

Carefully evaluate the table with matches to ensure that matching scores are acceptable and that most species have a corresponding gbif_id. Species that cannot be reliable matched or that don't have an associated gbif_id, can not contribute to the PD calculation and should be removed.

```{r}
matched <- matched %>% dplyr::filter(!is.na(gbif_id))
head(matched)
```

Then, we can use the function append_ott_id() to append the ott_id's as a new variable to the provided datacube, by joining on gbif_id.

```{r}
mcube <- append_ott_id(tree, cube, matched)
head(mcube)
```

When species in the datacube are not included in the provided phylogenetic tree, the ott_id variable will be *NA*. We can run the function check_completeness() to see how complete the provided phylogenetic tree is.
```{r}
mcube_dist <- distinct(mcube, speciesKey, .keep_all=TRUE)
head(mcube_dist)
```

We can check if the provided phylogenetic tree includes all species in the occurence cube by running the check_completeness() function on *mcube*.

```{r}
check_completeness(mcube)
```
Please note that occurence records for species that are not part of the provided phylogenetic tree will need to be removed - in case this number is large, please consider searching for a more complete phylogenetic tree that covers all your species!

```{r}
mcube <- mcube %>% dplyr::filter(!is.na(ott_id))
head(mcube)
```

### Calculate Phylogenetic Diversity for each grid cell

We first have to aggregate the occurrence cube in order to get a list of observed species for each grid cell. 

```{r}
aggr_cube <- aggregate_cube(mcube)
print(aggr_cube)
print(aggr_cube[[294,3]])
```

We can then calculate the PD value for all grid cells by using the purrr:map function to apply the function calculate_PD() for each grid cell.

PD_cube <- aggr_cube %>% mutate(PD = purrr::map(unique_names, calculate_pd, tree=tree))



### Visualize PD on a map & calculate indicator

Finally we read in the EEA Grid shapefiles and merge them to occurrence cube by joining on the eeaCellCode field.
The PD cube can then be plotted and overlayn with shapefile of WDPA protected areas shapefile.
)

<img src="./output/map_BE.pdf" alt="Map phylogenetic diversity of Musceloidea in Belgium"  width="4200" height="4200">
